---
title: "TOC and DOC Analysis"
author: "Kai Oda"
date: "11/2/2020"
output: github_document
---

## Loading in the packages and data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)

```

Now, lets read in the two excel sheets

```{r}
excel_sheets("../Input_Data/week4/144L_2018_Exp_TOC.xlsx")

#Read in the metadata sheet
metadata <- read_excel("../Input_Data/week4/144L_2018_Exp_TOC.xlsx", sheet = "Metadata")

#Read in the data sheet
data <- read_excel("../Input_Data/week4/144L_2018_Exp_TOC.xlsx", sheet = "Data")

#Join the two datasets together 
joined <- left_join(metadata, data)

str(joined)

```

After reading in our data, we want to perform some basic data wrangling to create a "days" column by treatment and bottle  We also want to subset the data to only include our columns of interest. 

```{r}
toc <- joined %>% 
  mutate(Datetime = ymd_hm(Datetime)) %>% 
  group_by(Treatment, Bottle) %>%
  mutate(interv = interval(first(Datetime), Datetime), 
         dur = as.duration(interv), 
         days= as.numeric(dur, "days")) %>% 
  ungroup() %>% 
  rename(sd_TOC = TOC_sd, 
         sd_PTOC = PTOC_sd) %>% 
  select(everything(), -c(contains("Sample"), interv, dur))
  

str(toc)
```

*Note: For some reason, the interval() function does not behave as it does in the videos. It returns two dates. I had to use the as.duration() function to calculate the time between the dates, then the as.numeric() function to convert that time into days.*

## Plotting TOC and DOC 

It is beneficial to plot TOC and DOC on the same graph to visually check for major differences. 

```{r}
pivot_toc <- toc %>% 
  select(Treatment, Bottle, days, TOC, PTOC)

#reduce rows and increase columns to have the values of TOC and PTOC in the same column
pivot_toc <- pivot_toc %>% 
  pivot_longer(TOC:PTOC, names_to="sample", values_to = "value")

pivot_toc_sd <- toc %>% 
  select(Treatment, Bottle, days, sd_TOC, sd_PTOC)  

## Do thte same thing for the sd
pivot_toc_sd <- pivot_toc_sd %>% 
  pivot_longer(sd_TOC:sd_PTOC, names_to="sample", names_prefix = "sd_", values_to = "sd")

## Combine the datsets and make the sample column easier to understand
pivoted <- left_join(pivot_toc, pivot_toc_sd) %>% 
  mutate(sample = ifelse(sample=="TOC", "Bottle", "Vial"))

## Set the order of variables for future graphs 
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

pivoted$Treatment <- factor(pivoted$Treatment, levels = levels)
pivoted$sample <- factor(pivoted$sample)
```

Now that our dataset is prepared we can begin plotting!

```{r fig.height=6}
pivoted %>% 
  drop_na(value) %>% 
  ggplot(aes(x=days, y=value, color=sample), group = interaction(Treatment, sample))+
  geom_errorbar(aes(ymin=value-sd, ymax = value+sd), width=0.1)+
  geom_point(shape=21, alpha=0.7, size=3)+
  labs(x="Days", y=expression("Total Organic Carbon µmol L"^-1))+
  facet_grid(rows = "Treatment")
  

```
From a glance it appears that TOC measurements between the bottle and vials are relatively similar. However, to ensure that they are statistically comparable a more robust test is needed. 


## Model II Linear Regression

Next, we want to compare the bottle and vial samples to determine whether the measurements between the two are statistically similar. 

A Model II linear regression assumes that both x and y are dependent variables that are measured and have error. In this case, x will represent TOC while y represents PTOC. 

```{r}
library(lmodel2)

## Drop na values and only include results after timepoint 0
toc_lm_data <- toc %>% 
  drop_na(TOC) %>% 
  filter(Timepoint > 0)

toc_lm <- lmodel2(PTOC ~ TOC, data=toc_lm_data, nperm=99)

## Now, we save the values of intercept and slope. 
intercept <- toc_lm$regression.results[3,2]
slope <- toc_lm$regression.results[3,3]

## Now, we save the values of the CI for intercept and slope
two_int <- toc_lm$confidence.intervals[3,2]
two_slope <- toc_lm$confidence.intervals[3,4]

nine_int <- toc_lm$confidence.intervals[3,3]
nine_slope <- toc_lm$confidence.intervals[3,5]

```

Now, lets plot the TOC and PTOC values in ggplot 

```{r}
ggplot(toc_lm_data, aes(x=TOC, y=PTOC))+
  geom_errorbar(aes(ymin=PTOC-sd_PTOC, ymax = PTOC + sd_PTOC), width=0.2)+
  geom_point(shape=21, size=3, fill="white", alpha=0.6)+
  geom_abline(intercept=intercept, slope=slope, linetype=2)+
  geom_abline(intercept=two_int, slope=two_slope, linetype=2)+
  geom_abline(intercept=nine_int, slope=nine_slope, linetype=2)+
  labs(x=expression("Bottle TOC µmol L"^-1), y = expression("Vial TOC µmol L"^-1))+
  annotate(geom="text", label = expression(atop("y = 1.03x - 2.1", paste("r"^2," = 0.96, ", italic("p "), "= 0.01"))), x = 80, y=90, size=4)
```

Most values can be explained by the regression line, p value is less than 0.05. I take this to mean that the measurements between vials and bottles were relatively similar. Therefore, this test us that the vial measurements are representative of the TOC concentrations from the large bottle. Thus, we can proceed with calculating DOC from the TOC values from the vials. 

## Calculating the DOC 

To calculate the DOC = TOC - POC. Unfortunately, we need bactierial carbon for POC. So let's get to it! 

```{r}
bc <- read_rds("../Bacterial_Abundance/bacterialCarbon.RDS")
merge <- left_join(toc, bc)

## In the next chunk we reorder the variabls so that time dependent variables appear first. Values that are not time dependent follow these time dependent variables 
merge <- merge %>% 
  select(Experiment:days, TOC:sd_PTOC, cells_L:delta_cells, bc, avg_bc, sd_bc, everything(), -c(contains("Sample"))) %>% 
  group_by(Treatment, Bottle) %>% 
  fill(exp_start:avg_lag_phase_dur) %>% 
  ungroup()
## We also fill in some missing values 

glimpse(merge)

```

Bactieral abundance/TOC from vial samples were not taken at every timepoint. In some cases we have a TOC sample but no bacterial carbon sample (or vice versa). We want to use an interpolation function to fill in the missing values between two known values for TOC and bacterial carbon (POC) so that we can compute DOC for a greater number of timepoints. 

```{r}
## Select the variables of interest...
to_interpolate <- merge %>% 
  select(Experiment, Treatment, Bottle, Timepoint, days, PTOC, bc) %>% 
  group_by(Treatment, Bottle)

## Use the group_split() function to break the dataset into a list (data type)
## The list is split up by the grouped elements (Treatment, Bottle)
list <- to_interpolate %>% 
  group_split()

## Generate the keys, (Treatment, Bottle)
keys <- to_interpolate %>% 
  group_keys() %>% 
  mutate(key = paste(Treatment, Bottle))

## Change the name of the list elements to the generated key values 
names(list) <- keys$key

```

### Writing the function

```{r}
library(zoo)

## Interpolates the toc and POC values, joins them together with cbind() and converts the matrix to a tbl with as_tibble()
interp.funct <- function(x) {
  y <- zoo(x, order.by = x$days)
  interp_toc <- round(as.numeric(na.approx(y$PTOC, na.rm=F)), 1)
  interp_bc <- round(as.numeric(na.approx(y$bc, na.rm=F)), 1)
  z <- cbind(y, interp_toc, interp_bc)
  as_tibble(z)
}

```

## Apply the function and combine the columns into the main dataset

```{r}
## lapply() takes a list and applies a function to each element in the list 
## We also need to convert numeric variables (they are chrs by default from as_tibble())
interpolated <- lapply(list, interp.funct) %>% 
  plyr::ldply(., as.data.frame) %>% 
  select(-c(.id, PTOC, bc, days)) %>% 
  mutate_at(vars(Timepoint:interp_bc), as.numeric) %>% 
  left_join(merge, .)

## Now we can see the MASSIVE dataframe with the interpolated values :) 
glimpse(interpolated)
```

## Calculating Derived Variables 

Now that we have our interpolated TOC and POC values we want to calculate a bunch of values related to DOC and BGE (bacterial growth efficiency). 

```{r}
doc <- interpolated %>% 
  mutate(doc = interp_toc - interp_bc) %>% 
  group_by(Treatment, Bottle) %>% 
  mutate(
    bioav_doc = (first(doc) - last(na.omit(doc))) / first(doc), 
    doc_exp_end = ifelse(Timepoint == exp_end, doc, NA), 
    delta_doc = first(doc) - doc_exp_end, 
    tdelta_doc = first(doc) - last(na.omit(doc)), 
    bge = ifelse(delta_doc > 1.5, deltaBc/delta_doc, NA)
  ) %>% 
  fill(bioav_doc:bge, .direction = "downup") %>% 
  ungroup()

## Don't forget to fill the columns!

subset <- doc %>% 
  select(Treatment, Bottle, Timepoint, doc, bioav_doc, doc_exp_end, delta_doc, tdelta_doc, bge)

str(doc)

```

## Plotting average derived values 

We've calculated DOC and a bunch of other variables. It is beneficial to see the averaged values across treatments and timepoints where applicable. First, we need to generate a datset containing the derived variables. 

```{r}
averages <- doc %>% 
  group_by(Treatment, Timepoint) %>% 
  mutate(avg_toc = mean(PTOC), 
         sd_toc = sd(PTOC)) %>% 
  ungroup() %>% 
  group_by(Treatment) %>%
  mutate(
    avg_bioav_doc = mean(bioav_doc),
    sd_bioav_doc = sd(bioav_doc),
    avg_delta_doc = mean(delta_doc),
    sd_delta_doc = sd(delta_doc),
    avg_tdelta_doc = mean(tdelta_doc),
    sd_tdelta_doc = sd(tdelta_doc),
    avg_bge = mean(bge),
    sd_bge = sd(bge)
  ) %>% 
  ungroup()

```


We have calculated the average values and added them to our giant dataset. Now, we want to plot these average values. 

## Plotting average values with ggplot 

First, lets plot the average TOC values for each treatment

```{r}
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

averages$Treatment <- factor(averages$Treatment, levels = levels)

averages %>% 
  mutate(Treatment = factor(Treatment)) %>% 
  drop_na(avg_toc) %>% 
  ggplot(aes(x=days, y = avg_toc, color = Treatment), group = interaction(Treatment)) +
  geom_errorbar(aes(ymin=avg_toc - sd_toc, ymax = avg_toc + sd_toc), width = 0.2)+
  geom_point()+
  geom_line()+
  labs(x="Days", y = expression("Average Total Organic Carbon, µmol C L"^-1))+
  theme_bw()

```
The two treatments with the highest growth rate/bacterial abundance (ash leachate and GNP), experienced the greatest decrease in TOC over the course of the experiment. This likely indicates that they used the available carbon to increase their biomass. Mud leachate treatment started at around the same level but little C was actually taken up by the microbes. 


Now, we'll make bar plots for the remaining averaged variables 

```{r}
bar_data <- averages %>% 
  select(Treatment, avg_bioav_doc:sd_bge) %>% 
  distinct()
```


**Average bioavailable DOC**
```{r fig.width=9}
bioav_doc_graph <- bar_data %>% 
  ggplot(aes(x=Treatment, y=avg_bioav_doc))+
  geom_col(aes(fill=Treatment))+
  geom_errorbar(aes(ymin = avg_bioav_doc - sd_bioav_doc, ymax = avg_bioav_doc + sd_bioav_doc), width=0.3)+
  labs(y="Bioavailable DOC Fraction")+
  theme_bw()


bioav_doc_graph

```
It appears that ash leachate and GNP treatments had the highest bioavailable DOC fraction. This is in line with the increased growth rates and large decrease in TOC over the course of the experiment. Interestingly, mud leachate treatments had a lower bioavailable doc fraction when compared to the control. This seems to indicate that there is not much bioavailable carbon in mud leachate. 


**Average Change in DOC**
```{r fig.width=9}
delta_doc_graph <- bar_data %>% 
  ggplot(aes(x=Treatment, y=avg_delta_doc))+
  geom_col(aes(fill=Treatment))+
  geom_errorbar(aes(ymin = avg_delta_doc - sd_delta_doc, ymax = avg_delta_doc + sd_delta_doc), width=0.3)+
  labs(y=expression("Change in DOC (to stationary phase), µmol C L"^-1))+
  theme_bw()


delta_doc_graph

```
Ash leachate and GNP treatments have the highest change in DOC, seeming to indicate that they made the greatest use of the available carbon to increase their biomass. 

**Average Change in DOC (total)**
```{r fig.width=9}
tdelta_doc_graph <- bar_data %>% 
  ggplot(aes(x=Treatment, y=avg_tdelta_doc))+
  geom_col(aes(fill=Treatment))+
  geom_errorbar(aes(ymin = avg_tdelta_doc - sd_tdelta_doc, ymax = avg_tdelta_doc + sd_tdelta_doc), width=0.3)+
  labs(y=expression("Change in DOC (total), µmol C L"^-1))+
  theme_bw()


tdelta_doc_graph

```
**See above explanation, results are similar. 


**Average BGE (total)**
```{r fig.width=9}
bge_graph <- bar_data %>% 
  ggplot(aes(x=Treatment, y=avg_bge))+
  geom_col(aes(fill=Treatment))+
  geom_errorbar(aes(ymin = avg_bge - sd_bge, ymax = avg_bge + sd_bge), width=0.3)+
  labs(y=expression("Bacterial Growth Efficiency"))+
  theme_bw()


bge_graph

```
It seems odd that there is no BGE for control or mud leachate treatments. However, this has to do with the way we calcualted BGE. Since the detection limit is > 1.5, we assigned NA values if the change in DOC concentrations was less than that. Control and mud leachate treatments did not exceed this detection limit, thus no BGE values are available. With that out of the way, ash leachate has a slightly higher BGE, indicating that a greater percentage of nutrients were used to increase biomass in this treatment though a more robust statistical test is needed to say whether this is a significant increase over the GNP treatment. 


Now, we can display all the graphs at once with pathwork()

```{r fig.height=15, fig.width=9}
library(patchwork)

bioav_doc_graph / delta_doc_graph / tdelta_doc_graph / bge_graph + plot_annotation(title="Averaged Plots for Ash,Mud,GNP Additions", tag_levels="a")
```


```{r}
saveRDS(averages, "averaged_DOC_dataset.RDS")
```











