---
title: "Phyloseq Analysis"
author: "Kai Oda"
date: "11/12/2020"
output: github_document
---

```{r}
# BiocManager::install("phyloseq")
```

## Load in the necessary libraries 
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)
library(RColorBrewer)
```

## Import Data from weeks 4 and 5

```{r}
count.tab <- read_rds("../DaDa2/seqtab-nochimtaxa.rds") 
tax.tab <- read_rds("../DaDa2/taxa.rds") 
sample.tab <- read_rds("../Bacterial_Abundance/bacterialCarbon.RDS") %>% 
  drop_na(DNA_SampleID) %>% 
  column_to_rownames(var = "DNA_SampleID") 
```

## Create a phyloseq object 

Essentially, all we are doing is combining the three datasets together 

```{r}
OTU = otu_table(count.tab, taxa_are_rows = TRUE) 
TAX = tax_table(tax.tab)
SAM = sample_data(sample.tab)
ps = phyloseq(OTU,TAX,SAM) 
```

## Filtering out chloroplasts and mitochondria 

We only want bacteria in our phyloseq object, so let's remove chloroplast and mitochondria taxa with the subset_taxa() command 

```{r}
sub_ps <- ps %>%
  subset_taxa(Family  != "mitochondria" & Order  != "Chloroplast")
```

## Plotting distribution of read lengths 

```{r}
sample_sum_df <- data.frame(sum = sample_sums(sub_ps))

ggplot(sample_sum_df, aes(x=sum))+
  geom_histogram(color="black", fill="white", binwidth = 5000)+
  xlab("Read counts")+
  ylab("Occurences")+
  ggtitle("Distribution of sample sequencing depth")+
  theme_bw()

```

```{r}
## Outputs mean, min and max of the read counts from the dataframe
summary(sample_sum_df)
```

One thing to note from this output: the mean read length is SIGNIFICANTLY larger than the minimum read length (different from the videos). 

This means that when we rarefy the data there will be a massive effect due to subsampling. Essentially, we are taking a dataset that has an average of 28000+ in read length, and only randomly sampling 2064 sequences from it. This seems like a LOT of lost data but we push on. 

## Rarefying the data 

As I understand from the video series: 

- In order to calculate beta diversity we need to calculate pairwise comparisons between the data 
- Pairwise comparisons are influenced by read depth 
- Therefore, it is best practice to normalize/standardize read depth across samples 
- This standardization process is called "rarefying" data
- Essentially, we scale to the smallest library size and sub sample all sequences until we have a library equal to the smallest size... 

```{r}
ps_min <-  rarefy_even_depth(sub_ps, sample.size = min(sample_sums(sub_ps)))

mean(sample_sums(sub_ps))#28808.5
mean(sample_sums(ps_min)) #2064 
```

## Ordinating and visualizing beta diversity 

```{r}
## Calculating the stress level for the non-subsampled datset 
set.seed(1)
nmds <- ordinate(sub_ps, method = "NMDS",  distance = "bray") 
```

Stress level for the **nonsubsampled dataset is 0.076.** This is lower than the cutoff (0.2). Therefore, this is a reasonably good translation of multi-dimensional data to a 2D space. 

```{r}
set.seed(1)
nmds_min <- ordinate(ps_min, method = "NMDS",  distance = "bray") 
```

Stress level for the subsampled dataset is **0.08**. This means that the subsampled datset is also a good translation from multi-dimensional data to 2D space. 

## Plotting beta diversity 

```{r fig.width=12}
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

nmds.plot <- plot_ordination(sub_ps, nmds, title="NMDS Nonsubsampled")+
  geom_point(aes(fill=days, shape = factor(Treatment, levels = levels)), alpha = 0.6, stroke = 2, size = 4)+
  scale_shape_manual(values = c(21, 22, 23, 24))+
  scale_fill_gradient(low = "blue", high = "#d31f2a")+
  theme_bw()

#remove the point layer (removes the point within the point)
nmds.plot$layers <- nmds.plot$layers[-1]

nmds.plot + 
  facet_grid(~Treatment)+
  guides(fill = guide_colorbar(title="Days"), shape= guide_legend(title="Treatment"))

```

```{r fig.width=12}
nmds_min.plot <- plot_ordination(ps_min, nmds_min, title="NMDS Subsampled")+
  geom_point(aes(fill=days, shape = factor(Treatment, levels = levels)), alpha = 0.6, stroke = 2, size = 4)+
  scale_shape_manual(values = c(21, 22, 23, 24))+
  scale_fill_gradient(low = "blue", high = "#d31f2a")+
  theme_bw()

#remove the point layer (removes the point within the point)
nmds_min.plot$layers <- nmds_min.plot$layers[-1]

nmds_min.plot + 
  facet_grid(~Treatment)+
  guides(fill = guide_colorbar(title="Days"), shape= guide_legend(title="Treatment"))
```






