---
title: "Phyloseq Analysis"
author: "Kai Oda"
date: "11/12/2020"
output: github_document
---

```{r}
# BiocManager::install("phyloseq")
```

## Load in the necessary libraries 
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(phyloseq)
library(RColorBrewer)
```

## Import Data from weeks 4 and 5

```{r}
count.tab <- read_rds("../DaDa2/seqtab-nochimtaxa.rds") 
tax.tab <- read_rds("../DaDa2/taxa.rds") 
sample.tab <- read_rds("../Bacterial_Abundance/bacterialCarbon.RDS") %>% 
  drop_na(DNA_SampleID) %>% 
  column_to_rownames(var = "DNA_SampleID") 
```

## Create a phyloseq object 

Essentially, all we are doing is combining the three datasets together 

```{r}
OTU = otu_table(count.tab, taxa_are_rows = TRUE) 
TAX = tax_table(tax.tab)
SAM = sample_data(sample.tab)
ps = phyloseq(OTU,TAX,SAM) 
```

## Filtering out chloroplasts and mitochondria 

We only want bacteria in our phyloseq object, so let's remove chloroplast and mitochondria taxa with the subset_taxa() command 

```{r}
sub_ps <- ps %>%
  subset_taxa(Family  != "mitochondria" & Order  != "Chloroplast")
```

## Plotting distribution of read lengths 

```{r}
sample_sum_df <- data.frame(sum = sample_sums(sub_ps))

ggplot(sample_sum_df, aes(x=sum))+
  geom_histogram(color="black", fill="white", binwidth = 5000)+
  xlab("Read counts")+
  ylab("Occurences")+
  ggtitle("Distribution of sample sequencing depth")+
  theme_bw()

```

From the histogram we can see that read counts are more or less normally distributed with a mean ~28000. However, there are some outliers on either side. The disparity between the smallest read count and the median may pose some issues when calculating beta diversity. 

```{r}
## Outputs mean, min and max of the read counts from the dataframe
summary(sample_sum_df)
```

One thing to note from this output: the mean read length is SIGNIFICANTLY larger than the minimum read length (different from the videos). 

This means that when we rarefy the data there will be a massive effect due to subsampling. Essentially, we are taking a dataset that has an average of 28000+ in read length, and only randomly sampling 2064 sequences from it. This seems like a LOT of lost data but we push on. 

## Rarefying the data 

As I understand from the video series: 

- In order to calculate beta diversity we need to calculate pairwise comparisons between the data 
- Pairwise comparisons are influenced by read depth 
- Therefore, it is best practice to normalize/standardize read depth across samples 
- This standardization process is called "rarefying" data
- Essentially, we scale to the smallest library size and sub sample all sequences until we have a library equal to the smallest size... 

```{r}
ps_min <-  rarefy_even_depth(sub_ps, sample.size = min(sample_sums(sub_ps)))

mean(sample_sums(sub_ps))#28808.5
mean(sample_sums(ps_min)) #2064 
```

## Ordinating and visualizing beta diversity 

```{r}
## Calculating the stress level for the non-subsampled datset 
set.seed(1)
nmds <- ordinate(sub_ps, method = "NMDS",  distance = "bray") 
```

Stress level for the **nonsubsampled dataset is 0.076.** This is lower than the cutoff (0.2). Therefore, this is a reasonably good translation of multi-dimensional data to a 2D space. 

```{r}
set.seed(1)
nmds_min <- ordinate(ps_min, method = "NMDS",  distance = "bray") 
```

Stress level for the subsampled dataset is **0.08**. This means that the subsampled datset is also a good translation from multi-dimensional data to 2D space. 

## Plotting beta diversity 

```{r fig.width=12}
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

nmds.plot <- plot_ordination(sub_ps, nmds, title="NMDS Nonsubsampled")+
  geom_point(aes(fill=days, shape = factor(Treatment, levels = levels)), alpha = 0.6, stroke = 2, size = 4)+
  scale_shape_manual(values = c(21, 22, 23, 24))+
  scale_fill_gradient(low = "blue", high = "#d31f2a")+
  theme_bw()

#remove the point layer (removes the point within the point)
nmds.plot$layers <- nmds.plot$layers[-1]

nmds.plot + 
  facet_grid(~Location)+
  guides(fill = guide_colorbar(title="Days"), shape= guide_legend(title="Treatment"))

```

Clustered points indicate similar diversity while distance points indicate relative disimilarity. In early timepoints all treatments are relatively similar. However, in later treatments the diversity diverges, with GNP and ash leachate treatments having distinct diversity. Control and mud leachate treatments appear to have similar diversity. 


```{r fig.width=12}
nmds_min.plot <- plot_ordination(ps_min, nmds_min, title="NMDS Subsampled")+
  geom_point(aes(fill=days, shape = factor(Treatment, levels = levels)), alpha = 0.6, stroke = 2, size = 4)+
  scale_shape_manual(values = c(21, 22, 23, 24))+
  scale_fill_gradient(low = "blue", high = "#d31f2a")+
  theme_bw()

#remove the point layer (removes the point within the point)
nmds_min.plot$layers <- nmds_min.plot$layers[-1]

nmds_min.plot + 
  facet_grid(~Location)+
  guides(fill = guide_colorbar(title="Days"), shape= guide_legend(title="Treatment"))
```

The subsampled beta diversity plot has similar trends to the nonsubsampled one. 

## Calculating and plotting alpha diversity 

We will be using two different calculations of alpha diversity: Chao1 and Shannon index 

- The Chao1 measure computes species richness 
- The Shannon index computers species richness **and** eveness 

```{r}
## Calculate alpha diversity measures 
richness <- estimate_richness(ps_min, measures = c("Chao1", "Shannon"))

## Create a column that contains the name of the rows
richness <- richness %>% 
  rownames_to_column(., var = "DNA_ID") %>% 
  mutate_at(vars(DNA_ID), str_replace_all, pattern = "X144", "144") 
## Using the mutate_at() function to replace X144 strings with 144 so that we can combine the metadata and richness estimates 

```


```{r}
# Join the alpha diversity measures to the metadata 
alphadiv <- left_join(richness, sample.tab %>% rownames_to_column(., var = "DNA_ID")) 
```

In the next step we want to use the pivot_longer() function to combine the alpha diversity measures into a single column. This will allow us to plot both measures on the same graph. We also compute the standard error associated with the Chao1 measure 

```{r}
pivot.data <- alphadiv %>% 
  select(Treatment, Bottle, Timepoint, days, Chao1, Shannon) %>% 
  pivot_longer(.,  cols = c(Chao1,  Shannon), names_to = "measure", values_to = "est" ) %>% 
  left_join(., alphadiv %>% 
              select(Treatment, Bottle, Timepoint, days, se.chao1)) %>% 
  mutate(se.chao1 = ifelse(measure == "Chao1", se.chao1, NA)) 
```

## Plotting alpha diversity 

```{r fig.height=8, fig.width=12, warning=FALSE}
library(ggpubr)

alpha.plot <- ggboxplot(pivot.data, x = "Timepoint", y = "est",
            xlab = expression(italic(paste(""))), 
            ylab = expression(italic(paste("Alpha Diversity Measure"))),
            add = "dotplot",
            width = 0.2,
            ggtheme = theme_bw()) +  
  stat_compare_means(label.x = "4") +
  facet_grid(measure~ factor(Treatment, levels = levels), scales = "free") 

alpha.plot
```
There does not seem to be a significant difference in the alpha diversity within treatments cross the experiment. This is somewhat confusing, as some treatments appear to have drastically different diversity depending on the timepoint. However, the Kruskal-Wallis test (similar to a one way ANOVA), indicates that there is no significant difference in the mean alpha diversity across time within treatments... 

```{r fig.height=8, fig.width=15}
alpha.plot2 <-  ggboxplot(pivot.data, x = "Treatment", y = "est", 
            xlab = expression(italic(paste(""))), 
            ylab = expression(italic(paste("Alpha Diversity Measure"))),
            add = "dotplot",
            width = 0.2,
            ggtheme = theme_bw()) +  
  stat_compare_means(label.x = "Mud Leachate") +
  facet_grid(measure~Timepoint, scales = "free") 

alpha.plot2

```

Once again, no significant difference in the mean alpha diversity across treatments. Generally, as time increases the p value decreases, but at no time point are the means significantly different from one another across treatments. 

## Creating a heat map 

We have calculated and visualized alpha and beta diversity, but now we want to learn a bit more about the taxa that influenced community composition over time. Specifically, which taxa increased or decreased over time across different treatments? 

Unfortunately, to answer that question we have to do a LOT of data manipulation :(

```{r}
# First, calculate the fraction of an ASV compared to the total number of ASVs 
ps_std <- transform_sample_counts(ps_min, function(x) x/sum(x))

# Get the relative abundance data as a matrix 
ps_std.tab <- as(otu_table(ps_std), "matrix")

# Converting the matrix relative abundance data to a dataframe 
ps_std.df = as.data.frame(ps_std.tab) 
```

We'll prepare a massive table that contains the relative abundance and taxonomic information 

```{r}
# Convert the taxa table to a dataframe
tax.df <-  as.data.frame(tax.tab) 

# Combine the taxa and relative abundance dataframes 
custom.tab <- tax.df %>% 
  rownames_to_column(., var = "asv") %>% 
  left_join(., ps_std.df %>% rownames_to_column(., var = "asv")) %>% 
#In order to simplify the data frame we combine the phylum, class and order columns into a new column that contains all three values 
  mutate(pco = paste(Phylum, "_", Class, "_", Order)) %>% 
## Remove unecessary data 
  select(-c(asv:Genus)) %>% 
## Group data frame by unique phylum, class and orders 
  select(pco,everything()) %>% 
  group_by(pco) %>% 
## Combine relative abundances of the same pco 
  summarise_at(vars(contains(c("144"))), sum, na.rm = T) %>% 
  ungroup()
```

To make things a bit easier for us (to look through the data) we want to have our taxonomic groups as column names and sample numbers as row names. We will transpose the data to make this happen 

```{r}
## We save the row names of the dataframe into a variable for later use 
colnames <- custom.tab[,1] 

## Transposing the data (turning the rows into columns and vice versa)
t_custom.tab <-  as.data.frame(t(custom.tab[,-1]))

## And then we assign the column names 
colnames(t_custom.tab) <- colnames$pco

# We prepare the dataframe that contains our metadata information
## Convert the rows names to a column name and select the necessary variables 
sample.tab <- sample.tab %>% 
  rownames_to_column(., var="sample") %>% 
  select(sample, Experiment, Location, Bottle, Treatment, Timepoint, days, cells_L)

## Then, we merge the dataframe to the massive metadata dataframe to get timepoint, treatment, bottle etc. information associated with the relative abundance and taxa data 
sweet.tab <- t_custom.tab %>% 
  rownames_to_column(., var = "sample") %>% 
  left_join(., sample.tab) %>% 
  select(sample, Experiment:cells_L, everything())
```

I called the "cells" column **cells_L** all the way back in the week 3/4 videos and that really came back to haunt me :o 

Some more data cleaning 

```{r}
relabund <- sweet.tab %>% 
  select(-c(sample:cells_L)) %>% 
# remove the taxa that are not present in any treatment (make the dataset more manageable)
  .[ , colSums(.) > 0] %>% 
# arrange the dataset by the biggest contributers
  .[, order(colSums(-.))] %>% 
  bind_cols(sweet.tab %>% select(sample:cells_L), .)
```

SO NOW WE HAVE A GREAT LOOKING DATASET AND WE CAN PLOT THE HEAT MAP ::::)))))

```{r}
relaheat.data <- relabund %>% 
## Remove unecessary columns
  select(-c(sample, Experiment, Location, Bottle, days, cells_L)) %>%
## Combine taxa information and relative abundance into a single column 
  pivot_longer(.,-c(Treatment:Timepoint), names_to = "taxa", values_to = "relabund") %>% 
## Separate our phylum, class, order column into distinct columns
  separate(taxa, into = c("p", "c", "o"), sep = " _ ") %>% 
## Group by the treatment, timepoint, phylum, class and order
  group_by(Treatment, Timepoint, p, c, o) %>% 
## Create a new column that is the average relative abundance for the previous grouping
  mutate(mean_relabund = mean(relabund, na.rm = T)) %>% 
  ungroup() %>% 
## Remove the relative abundance column (we will use the mean relative abundance from now on)
  select(-relabund) %>% 
## Only keep distinct values 
  distinct()
```

## Plotting the heat map 

```{r fig.height=12, fig.width=12}
## Import the color library we will be using 
library(viridis)

relaheat <- relaheat.data %>%
  ggplot(aes(x = Timepoint, y = o)) +
  geom_tile(aes(fill = mean_relabund), color = "white") +
  scale_fill_viridis(option = "C") +
  labs(x = "Timepoint", y = "Order", fill = "Relative Abundance") +
facet_grid(~factor(Treatment, levels = levels)) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 12),
        legend.position = "top") +
   guides(fill = guide_colourbar(barheight = 2, barwidth = 20, frame.colour = "black", frame.linewidth = 2,ticks.colour = "black", ticks.linewidth = 1), color = F) 

relaheat
```

General trends can be observed by simply looking at the heat plot. Oceanospirillales seemed to be much more prevalent in GNP treatments while Campylobacterales saw a surge in ash leachate treatments. 


## Saving our data

```{r}
saveRDS(sweet.tab, "Custom_ASV_Table.rds")
saveRDS(sub_ps, "phyloseq_obj.rds")
saveRDS(ps_min, "subsampled_phyloseq_obj.rds")
saveRDS(alphadiv, "alphadiv.rds")
```








