---
title: "Bacterial Abundance Analysis"
author: "Kai Oda"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)
library(dplyr)
```

## Loading in the data 

```{r}
# First let's see what sheets are within the xcel file that we will be working with 
excel_sheets("../Input_Data/week3/144L_2018_BactAbund.xlsx")

#Import the metadata 
metadata <- read_excel("../Input_Data/week3/144L_2018_BactAbund.xlsx", sheet="Metadata")

#unique treatments 
unique(metadata$Treatment)
```

Properties of this dataset
- Only **1 location** and one depth is sampled 
- There are **8 bottles** labeled (A-H)
- There are **9 timepoints** taken for each bottle 
- There are **4 treatments.** Control, ash leachate, mud leachate and the GNP mix. 
- There are **3 DNA samples** taken per bottle. 

```{r}
# Import the actual data 
data <- read_excel("../Input_Data/week3/144L_2018_BactAbund.xlsx", sheet="Data")

# Now, we join the actual data to the metadata 
joined <- left_join(metadata, data)

```

So now we have a dataset that contains the number of cells. *left_join()* synchronzied Cells_ml by Bottle and Timepoint 

## Data exploration/basic manipulation 

```{r}
str(joined)
```

We need to: 

- Convert cells/mL -> cells/L 
- Calculate duration between timepoints based on treatment and bottle 
- remove rows with NA values for cells 

``` {r}
cells <- joined %>% 
  mutate(Datetime = ymd_hm(Datetime), 
         cells_L = Cells_ml * 1000) %>% 
  drop_na(cells_L) %>% 
  group_by(Treatment, Bottle) %>% 
  mutate(interv = interval(first(Datetime), Datetime), 
         dur = as.duration(interv),
         days = as.numeric(dur, "days")) %>% 
  mutate(cells_sd = sd(cells_L)) %>% 
  ungroup() 
  

str(cells)
```


## Basic data visualization 
Normally, we could create a variable to store the order in which elements should appear. However, I am happy with the default ordering. 

```{r}
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

# This line orders the treatment variable by the specified order for all future figures
cells$Treatment <- factor(cells$Treatment, levels = levels)

cells %>% 
  mutate(dna = ifelse(DNA_Sample == TRUE, "*", NA)) %>% 
  ggplot(aes(x=days, y=cells_L, color=Treatment, 
             group = interaction(Treatment, Bottle)))+
  geom_point(size=3)+
  geom_errorbar(aes(ymin = cells_L - cells_sd, ymax = cells_L + cells_sd), width=0.1)+
  labs(x = "Days", y = expression(paste("Cells L"^-1)))+
  geom_line()+
  facet_grid(row="Treatment", scales="free")+
  geom_text(aes(label=dna), size=12, color="black")+
  theme_bw()


```

For the next part we need to plot the ln(cell_abundance) vs time (days). We will visually estimate where graph is linear. The slope of this linear portion will give us the specific growth rate. 

```{r}
ln_cells <- cells %>% 
  group_by(Treatment, Bottle) %>% 
  mutate(ln_cells = log(cells_L), 
         diff_ln_cells = ln_cells-lag(ln_cells, default=first(ln_cells))) %>% 
  ungroup()
```

Once we have calculated the difference in the natural log of the cell counts we can just plot the same graph and visually check for where the lines appear linear. 

```{r}
# Trying to display all 4 treatments and bottles in one graph makes it impossible to distinguish the linear portion
## Therefore, I am trying to split the dataset into two treatments each to make it easier to visualize 
ln_control_ash <- ln_cells %>% 
  filter(Treatment %in% c("Control", "Ash Leachate"))

ln_control_ash %>% 
  mutate(dna = ifelse(DNA_Sample == TRUE, "*", NA)) %>% 
  ggplot(aes(x=days, y=diff_ln_cells, color=Treatment, 
             group = interaction(Treatment, Bottle)))+
  geom_point(size=3)+
  labs(x = "Days", y = expression(paste("Change in ln(Cells) L"^-1)), title="Change in ln(Cells) in control + ash treatments over time")+
  geom_line()+
  facet_grid(Bottle~Treatment, scales="free")+
  geom_text(aes(label=dna), size=12, color="black")+
  theme_bw()

# Second graph, showing mud and GNP solution 
ln_mud_GNP <- ln_cells %>% 
  filter(Treatment %in% c("Mud Leachate", "Glucose_Nitrate_Phosphate"))

ln_mud_GNP %>% 
  mutate(dna = ifelse(DNA_Sample == TRUE, "*", NA)) %>% 
  ggplot(aes(x=days, y=diff_ln_cells, color=Treatment, 
             group = interaction(Treatment, Bottle)))+
  geom_point(size=3)+
  labs(x = "Days", y = expression(paste("Change in ln(Cells) L"^-1)), title="Change in ln(Cells) in mud + GNP treatments over time")+
  geom_line()+
  facet_grid(Bottle~Treatment, scales="free")+
  geom_text(aes(label=dna), size=12, color="black")+
  theme_bw()

```

Based on the graphs the we can visually estimate where exponential growth occurs for each bottle: 

- Control, Bottle A, ~0-0.5d (T0-T1)
- Control, Bottle B, ~0-0.5d (T0-T1)
- Ash Leachate, Bottle C, ~0-0.5d (T0-T1)
- Ash Leachate, Bottle D, ~0-0.5d (T0-T1)
- Mud Leachate, Bottle E, ~0-0.5d (T0-T1)
- Mud Leachate, Bottle F, ~0-0.5d (T0-T1)

- GNP, Bottle G, ~0-1d (T0-T2)
- GNP, Bottle H, ~0-1d (T0-T2)

## Calculate growth rates, doubling times and change in cell abundances 

```{r}
growth <- ln_cells %>% 
  mutate(exp_start = 0, 
         exp_end = ifelse(Treatment=="Glucose_Nitrate_Phosphate", 2, 1)) %>% 
  group_by(Treatment, Bottle) %>% 
  mutate(ln_cells_exp_start = ifelse(exp_start == Timepoint, ln_cells, NA), 
         ln_cells_exp_end = ifelse(exp_end == Timepoint, ln_cells, NA), 
         cells_exp_start = ifelse(exp_start == Timepoint, cells_L, NA), 
         cells_exp_end = ifelse(exp_end == Timepoint, cells_L, NA), 
         days_exp_start = ifelse(exp_start == Timepoint, days, NA), 
         days_exp_end = ifelse(exp_end == Timepoint, days, NA),) %>%
  fill(ln_cells_exp_start:days_exp_end, .direction="downup") %>%
  fill(ln_cells_exp_end, .direction="updown") %>% 
  mutate(mew = (ln_cells_exp_end - ln_cells_exp_start)/(days_exp_end - days_exp_start), 
         doubling_time = log(2)/mew, 
         delta_cells = cells_exp_end - first(cells_L)) %>% 
  ungroup() 


check <- growth %>%  
  select(Treatment, Bottle, Timepoint, ln_cells, exp_start, exp_end, cells_exp_start, cells_exp_end, mew, doubling_time)

```


## Calculate Bacterical Carbon 

- Convert bacterial abundance and change in bacterial abundance to C units 
- Apply a C conversion factor to bacterial abundnace (cells/L)
- This generates bacterial carbon (µmol C/L)

We will apply C content of bacterioplankton cells from Coastal Japan (-10 fg C/cell)

- Fukuda et. al., 1998

```{r}
batCarbon <- growth %>% 
  mutate(bc = cells_L * (2.5 * 10^-9), 
         deltaBc = delta_cells * (2.5*10^-9))

str(batCarbon)

```

## Calculating Treatment Averages 

Now that we have verified that the data from each bottle seems valid, it can be helpful to calculate the average values for important properties. In the following code chunk we will calculate the average values and standard deviation for: 

- Bacterial Carbon/cell 
- Specific growth rate 
- Doubling time 
- Change in cells over experiment 
- Change in cell C over experiment 

As a bonus we can also calculate the average duration of the lag phase over the course of the experiment 

```{r}
averages <- batCarbon %>% 
  group_by(Treatment, Timepoint) %>% 
  mutate(avg_bc = mean(bc), 
         sd_bc = sd(bc)) %>% 
  ungroup() %>% 
  group_by(Treatment) %>% 
  mutate(avg_mew = mean(mew), 
         sd_mew = sd(mew), 
         avg_doubling_time = mean(doubling_time), 
         sd_doubling_time = sd(doubling_time), 
         avg_delta_cells = mean(delta_cells), 
         sd_delta_cells = sd(delta_cells), 
         avg_delta_bc = mean(deltaBc), 
         sd_delta_bc = sd(deltaBc), 
         avg_lag_phase_dur = mean(days_exp_start)) %>% 
  ungroup()

subset <- averages %>% 
  select(Treatment, Timepoint, Bottle, bc, avg_bc, sd_bc)
  

```

Now that we have the dataset made with a bunch of average values we can make a plot!

First up, average carbon/cell 

```{r}
ggplot(averages, aes(x=days, y=avg_bc, color=Treatment))+
  geom_point()+
  geom_line()+
  geom_errorbar(aes(ymin= avg_bc - sd_bc, ymax = avg_bc + sd_bc), width=0.1)+
  facet_grid(row = "Treatment")+
  labs(x="Days", y=expression("Bacterial Carbon µmol C L"^-1))+
  theme_bw()

```

## Visualizing the rest of the data

We were able to plot average carbon/cell, but we have a bunch of other values! However, there is only one value for each treatment, so we will be making use of a new graph.... The column graph! 

```{r}
## First we need to subset the data to only include the variables of interest 
bar_data <- averages %>% 
  select(Location, Treatment, avg_mew:avg_lag_phase_dur)

## For further simplification we will only return unique values with the distinct() function 
bar_data <- bar_data %>% 
  distinct()

```

## Let's start plotting! 

```{r}
avg_mew <- ggplot(bar_data, aes(x=Treatment, y=avg_mew))+
  geom_col(fill="white", color="black")+
  geom_errorbar(aes(ymin=avg_mew-sd_mew, ymax = avg_mew + sd_mew), width=0.1)+
  labs(x="Treatment", y=expression("µ, d"^-1))+
  theme_bw()

avg_mew

```

```{r}
avg_doubling_time <- ggplot(bar_data, aes(x=Treatment, y=avg_doubling_time))+
  geom_col(fill="white", color="black")+
  geom_errorbar(aes(ymin=avg_doubling_time-sd_doubling_time, ymax = avg_doubling_time + sd_doubling_time), width=0.1)+
  labs(x="Treatment", y=expression("Doubling Time (days)"))+
  theme_bw()

avg_doubling_time

```


```{r}
avg_delta_bc <- ggplot(bar_data, aes(x=Treatment, y=avg_delta_bc))+
  geom_col(fill="white", color="black")+
  geom_errorbar(aes(ymin=avg_delta_bc-sd_delta_bc, ymax = avg_delta_bc + sd_delta_bc), width=0.1)+
  labs(x="Treatment", y=expression("Change in Bacterial Carbon, µmol C L"^-1))+
  theme_bw()

avg_delta_bc

```

```{r}
avg_lag_phase_dur <- ggplot(bar_data, aes(x=Treatment, y=avg_lag_phase_dur))+
  geom_col(fill="white", color="black")+
  labs(x="Treatment", y=expression("Average Duration of Lag Phase, days"))+
  theme_bw()

avg_lag_phase_dur

```


```{r fig.height=6, fig.width=13}
library(patchwork)

avg_mew + avg_delta_bc + avg_doubling_time + avg_lag_phase_dur + plot_annotation(tag_levels="a")

```












