---
title: "Thomas Fire Progression (Kai Oda)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
library(matrixStats)
library(readxl)
```

## Loading the data 
```{r}
fire_data <- read_excel("../Input_Data/week1/Thomas_Fire_Progression.xlsx", sheet="Data")
metadata <- read_excel("../Input_Data/week1/Thomas_Fire_Progression.xlsx", sheet="Metadata")
```

## Exploring the data
```{r}
glimpse(fire_data)
# It seems like the str() function is more useful bc it gives you the data type of each column 
str(fire_data)
#Look at first six rows
head(fire_data)
#Get the dimensions of the dataset
dim(fire_data)
## 39 rows, 5 columns 

```

## Manipulating the dataframe

Summarizing the data by columns 
```{r message=TRUE}
summary_data1 <- select(fire_data, c(Acres_Burned, Containment, PM10, PM25))

# Checking my dataset to see if there is na values...
sum_data <- summarise_all(summary_data1, ~sum(is.na(.)))

# Removing rows that contain na values 
summary_data1 <- na.omit(summary_data1)

# Calculating mean and variance for all numeric variables 
mean_data <- summarise_all(summary_data1, mean)
mean_data
var_data <- summarise_all(summary_data1, var)
var_data

## Practice renaming the column names
rename(summary_data1, acres_burned = Acres_Burned, containment = Containment, pm10 = PM10, pm25 = PM25)


```

## Plots 

**Plotting the acres burned by date**
```{r}
ggplot(fire_data, aes(x=Date, y=Acres_Burned))+
  geom_col()+
  labs(y="Acres Burned", title="Acres Burned throughout Thomas Fire")
```

This figure gives me an overview of how the Thomas Fire spread. We can see an exponential increase in acres burned in the first week, and then the destruction eventually leveling off. One interesting thing is that the "Acres_Burned" plummets down to a very low number write on Dec 12.  Perhaps this is an error in the dataset? 


**Plotting containment by date**
```{r}
ggplot(fire_data, aes(x=Date, y=Containment))+
  geom_col()+
  labs(y="Containment (%)", title="Containment of the Thomas Fire through time")
```

This graph simply gives us an idea of how long it took to contain the Thomas Fire. We can see that containment increase was mostly linear rather than exponential which is somewhat interesting. 

**Plotting particulate matter on same graph**
```{r}
#Restructuring the fire_data dataset so that both PM10 and PM25 can be displayed on the same graph 
## I assign both particulates to the same column, and create a new column that labels each value as either PM10 or PM25
pm_df <- fire_data %>%  
  select(Date, PM10, PM25) %>% 
  gather(key="type", value="amount", -Date)

ggplot(pm_df, aes(x=Date, y=amount, color=type))+
  geom_line()+
  labs(y="Particulate Matter (microns)", title="PM10 vs. PM25 from the Thomas Fire over time")

```

This graph shows how PM10 and PM25 vary throughout the Thomas Fire. Note the PM25 spike in the early days of the fire. It is also interesting that PM25 values fall sharply after that to track closely with PM10 values after. 
